---
title: "Test-Retest Reliability"
subtitle: "Simulating consistency, practice effects, and what correlation does (and doesn't) tell you"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
#    toc-title: "Contents"
    toc-location: left
    code-fold: true
    code-summary: "Show code"
#    embed-resources: true
    highlight-style: github
execute:
  warning: false
  message: false
  echo: false
---

## What is Test-Retest Reliability?

**Test-retest reliability** refers to the consistency of a measure when the same individuals are tested more than once under identical (or near-identical) conditions. It answers a deceptively simple question:

> *If I measure the same thing twice, do I get the same answer?*

A high test-retest correlation tells you the instrument is **consistent** — not that the construct itself is stable, or that the scores are accurate. This distinction matters more than it might seem, and we'll return to it at the end.

---

## Interactive Sandbox

The app below lets you change simulation parameters and see the results update in real time. It runs on a dedicated Shiny server — use the controls inside it to explore.

<style>
  iframe {
    display: block;
    box-sizing: border-box;
  }
</style>


<iframe src="https://app.hsancamillo.it/app/corr?sp_hide_navbar=true" style="border:none;" height="500" width="100%" title="Iframe Example"></iframe>



::: {.callout-tip title="Things to try in the app"}
- Increase **sample size** and watch the observed *r* stabilise toward the true ρ  
- Add a **practice effect** — the mean difference grows but *r* stays the same  
- Set **ρ = 0** and check that the t-test is non-significant  
:::

---

## Key Concepts

### Why does correlation ignore practice effects?

Pearson's $r$ measures the **linear relationship between deviations from the mean**.
When you add a constant $\Delta$ to every retest score:

$$r_{X,\, Y+\Delta} = \frac{\text{Cov}(X,\, Y+\Delta)}{\text{SD}(X)\cdot\text{SD}(Y+\Delta)} = \frac{\text{Cov}(X,Y)}{\text{SD}(X)\cdot\text{SD}(Y)} = r_{X,Y}$$

The constant cancels out entirely. The **paired t-test**, by contrast, works on
the raw differences $d_i = X_i - Y_i$, so it detects the shift immediately.

### Static R example

The chunk below replicates a single scenario in plain R — useful as a reference
for the formulas the Shiny app uses under the hood.

```{r}
#| label: static-example

library(faux)

set.seed(42)
res        <- rnorm_multi(50, mu = c(20, 20), sd = c(5, 5), r = 0.6)
names(res) <- c("test", "retest")

# correlation
cor.test(res$test, res$retest)

# paired t-test (no practice effect)
t.test(res$test, res$retest, paired = TRUE)

# paired t-test (practice effect +2)
t.test(res$test, res$retest + 2, paired = TRUE)
```

---

## Key Takeaway

::: {.callout-warning title="Don't confuse reliability with stability"}
A **high test-retest correlation** tells you that individuals who score high on
the first occasion tend to score high on the second — it reflects
**consistency of rank ordering**.

It does *not* tell you:

- That the absolute scores are accurate  
- That the construct hasn't changed over time  
- That systematic biases (like a practice effect) are absent  

For a fuller picture, pair the correlation with a check on mean-level
differences — which is exactly what **Bland-Altman plots** and **ICC** are
designed for.
:::

---

## Session Info

```{r}
#| label: session-info
#| code-fold: true
#| code-summary: "Show session info"

sessionInfo()
```
